---
title: Data wrangling in the tidyverse
subtitle: Using dplyr and tidy data 
author: Thomas de Graaff
date: April 21, 2021
institute: Department of Spatial Economics
header-includes:
- \usepackage{booktabs, array, multirow, dirtree}
- \usepackage{tikz, pgfplots}
- \pgfdeclarelayer{background}
- \pgfsetlayers{background,main}
output: binb::metropolis
fontsize: 12pt
---
  
  ```{r,setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Why bother?

1. For 20% of your time fancy techniques
    
2. For the other 80%
    - working with \alert{real} and \alert{messy} data
    - that has to be read in, cleaned, restructured, changed, described, communicated, visualised, etc.
    - in a \alert{structural, consistent \& reproducable} way
    
![](../figs/datascience)

## So why R and `tidyverse`?

> R was written by statisticians for statisticians

\begin{columns}
	\begin{column}{0.5\textwidth}
		\begin{itemize}
			\item \alert{multiple} ways in base R
			\item \texttt{dplyr} and \texttt{ggplot2}---later bundled in \texttt{tidyverse}
			\begin{itemize}
			  \item a more \alert{structured, encompassing, readable} approach to data wrangling
			 \end{itemize}
			\item \alert{many, many, many} offspring (\texttt{sf}, \texttt{dtplyr}, \texttt{ggraph})
		\end{itemize}
	\end{column}
		\begin{column}{0.5\textwidth}
		\begin{center}
			\includegraphics[width=0.9\textwidth]{../figs/wickham}
		\end{center}
	\end{column}
\end{columns}

## Invoking the `tidyverse`

```{r, warning = FALSE}
library("tidyverse")
```

## Read in data

```{r}
library("nycflights13")
fdata <- flights
```

display the first six rows of the dataframe `fdata`
```{r, eval = FALSE}
summary(fdata)
```
or view the dataset
```{r, eval = FALSE}
glimpse(fdata)
```

or look at the structure of the dataset
```{r, eval = FALSE}
str(fdata)
```

## Dataframes and tibbles

+ `fdata` is now a tibble (almost the same as a dataframe)

+ dataframes and tibbles

  - observations in rows
  - variables in columns (can be of different type)
  - a list of equal length vectors
  
## Example

```{r}  
grades <- tibble(
    name = c("Erik", "Eric", "Thomas"),
    quiz_1 = c(10, 9, 4),
    exam = c(7, 6, NA)
)
grades
```

# dplyr

## Why a need for `dplyr`

- Most data manipulation is done in Excel

    - sorting
    - creating/transforming variables
    - renaming variables
    - selecting variables/filtering observations

- But this, and more, can be done in `R` as well:

    - `dplyr` package 
    - big advantage: work can be \alert{reproduced}!
    - faster/scablable with bigger datasets
    - computations with \alert{groupings} of data
    - Cheatsheets to be found [here](https://www.rstudio.com/resources/cheatsheets/)
    
## `dplyr` basic verbs

- `dplyr` revolves around 5 verbs 

    - `arrange()`
    - `filter()` 
    - `select()`
    - `mutate()`
    - `summarize()`

- Other important commands (there are more):

    - `group_by()`
    - `rename()`
    - `distinct()`
    - count (`n()`)
    
- and can as well merge & restructure datasets

## You might want to sort variables

Sorting by arrival delay
```{r, warning=FALSE, message=FALSE, eval=FALSE}
arrange(fdata, arr_delay)
```  
In descending order
```{r, warning=FALSE, message=FALSE, eval=FALSE}
arrange(fdata, -arr_delay)
```  
Sorting by carrier arrival delay
```{r, warning=FALSE, message=FALSE, eval=FALSE}
arrange(fdata, carrier, -arr_delay)
``` 

## Filtering out observations

Only keep for observation on the first of May 
```{r, warning=FALSE, message=FALSE, eval=FALSE}
fdata_0501 <- filter(fdata, month == 5, day == 1)
```
Or remove those pesky missing values

```{r, warning=FALSE, message=FALSE, eval=FALSE}
filter(fdata, !is.na(arr_time))
```

## Selecting variables


Only select arrival delay and carrier
```{r, warning=FALSE, message=FALSE, eval=FALSE}
select(fdata, arr_delay, carrier)
```
Or select a \alert{range} of variables
```{r, warning=FALSE, message=FALSE, eval=FALSE}
select(fdata, month : arr_time )
```       

## Creating new variables

Perhaps you would like to have arrival delay squared
```{r, warning=FALSE, message=FALSE, eval=FALSE}
mutate(fdata, arr_delay_2 = arr_delay^2 )
``` 
Or change from minutes to seconds
```{r, warning=FALSE, message=FALSE, eval=FALSE}
mutate(fdata, arr_delay_s = arr_delay/60)
``` 

## And finally summarizing stuff (hey; statistics!)

Say, you would like to have the mean and standard deviation of arrival delay
```{r, warning=FALSE, message=FALSE}
summarize(fdata, 
          ave_arr_delay = mean(arr_delay),
          sd_arr_delay = mean(arr_delay)
          )
``` 

## Hey!

But not automatically corrected for missings!
```{r, warning=FALSE, message=FALSE}
summarize(fdata, 
          ave_arr_delay = mean(arr_delay, na.rm = TRUE),
          sd_arr_delay = sd(arr_delay, na.rm = TRUE)
)
```

## Other stuff

Rename stuff

```{r, warning=FALSE, message=FALSE, eval=FALSE}
rename(fdata, airline = carrier)
``` 
Remove duplicate rows:
```{r, warning=FALSE, message=FALSE, eval=FALSE}
distinct(fdata, carrier)
``` 
Count stuff 
```{r, warning=FALSE, message=FALSE, eval=FALSE}
summarize(fdata, count = n() )
``` 


## Making this work! Group data

```{r, warning=FALSE, message=FALSE}
fdata_carrier <- group_by(fdata, carrier)
summarize(fdata_carrier, 
          mean_arr_delay = mean(arr_delay, na.rm = TRUE),
          number = n())
``` 

## Getting jiggy with it

```{r, warning=FALSE, message=FALSE}
fdata_carrier_month <- group_by(fdata, carrier, month)
summarize(fdata_carrier_month, 
          mean_arr_delay = mean(arr_delay, na.rm = TRUE),
          number = n())
``` 

## Chaining your verbs: pipes

You do not need to create all kinds of new intermediate datasets. You can also link commands by using the pipe (`%>%`) operator:
```{r, warning=FALSE, message=FALSE}
fdata_carrier <- fdata %>%
            group_by(carrier) %>%
            summarize(
                mean_arr_delay = mean(arr_delay, 
                                      na.rm = TRUE),
                number = n())
```

## In combination with `ggplot2`

```{r, warning=FALSE, message=FALSE, eval=FALSE}
ggplot(fdata_carrier, 
       aes(x = reorder(carrier, -mean_arr_delay), 
           y = mean_arr_delay)) + 
    geom_bar(stat = "identity") + 
    theme_bw() + 
    labs(x = "Airline code", y = "Mean arrival delay")
```

## Or just go wild!

We want to plot arrival delay by month and carrier:
```{r, warning=FALSE, message=FALSE}
fdata_carrier_month <- fdata %>%
    group_by(carrier, month) %>%
    summarize(
        mean_arr_delay = mean(arr_delay, na.rm = TRUE),
        number = n())
```

## and plot it!
```{r, warning=FALSE, message=FALSE, eval=FALSE}
ggplot(fdata_carrier_month, 
       aes(x = carrier, y = mean_arr_delay)) + 
    geom_bar(stat = "identity") + 
    theme_bw() + 
    labs(x = "Airline code", 
         y = "Mean arrival delay") + 
    theme(text = element_text(size = 12) , 
          axis.text.x = element_text(angle=90, 
                                     hjust=1))  + 
    facet_wrap(~month)
```

# Tidy data

## Tidy data: what's that?

> Happy families are all alike; every unhappy family is unhappy in its own way — Leo Tolstoy

- A dataset is a collection of \alert{values}, usually either numbers (if quantitative) or strings (if qualitative)

  - every column is a variable
  - every row is an observation
  - every cell is a single value

- Values are organised in two ways. Every value belongs to a \alert{variable} (rows) and an \alert{observation} (columns).

## When do you need it?

- Requirements of programs or packages

  - for visualisation (`ggplot2` or `sf`)
  - for estimation (regression (?), conditional logits!, time-series, spatial statistics, `igraph`, etc.)
  
- Because you need to restructure data

  - for reading in data (remember sawtooth output!)
  - data from government websites (Statline!)
  
- Structure depends on \alert{unit} of analysis
  
## Violations of tidyness

- Column headers are values, not variable names

- Multiple variables are stored in one column

- Variables are stored in both rows and columns

- [Multiple types of observational units are stored in the same table]

- [A single observational unit is stored in multiple tables]

## Example


```{r, warning=FALSE, message=FALSE, eval=FALSE}
grades <- tibble(
    name = c("Erik", "Eric", "Thomas", "Jos"),
    quiz_1 = c(10, 9, 4, 8),
    quiz_2 = c(9, 9, 5, 7),
    exam = c(7, 6, NA, 7)
)
grades
```
Note: This is fine if you are interested in the whole educational career!

## Example (strikes back)

But if you are interested in grades, then invoke `pivot_longer`

```{r, warning=FALSE, message=FALSE, eval=FALSE}
grades2 <- grades %>% 
    pivot_longer(quiz_1:exam, 
                 names_to = "assessment", 
                 values_to = "grade") %>% 
    arrange(name, assessment)
grades2
```

## Similar example

Time-series context

```{r, warning=FALSE, message=FALSE, eval=FALSE}
billboard
billboard2 <- billboard %>% 
    pivot_longer(
        wk1:wk76, 
        names_to = "week", 
        values_to = "rank", 
        values_drop_na = TRUE
    )
billboard2
```

## Multiple values in one cell

```{r, warning=FALSE, message=FALSE, eval=FALSE}
who1 <- who %>% 
    pivot_longer(
        cols = new_sp_m014:newrel_f65, 
        names_to = "key", 
        values_to = "cases", 
        values_drop_na = TRUE
    )
who1
```

## Multiple values in one cell (revisited)

```{r, warning=FALSE, message=FALSE, eval=FALSE}
who2 <- who1 %>% 
    mutate(key = stringr::
             str_replace(key, "newrel", "new_rel")) >% 
    separate(key, c("new", "type", 
                    "sexage"), sep = "_") >% 
    select(-new, -iso2, -iso3) >% 
    separate(sexage, c("sex", "age"), sep = 1) >% 
who2
```

## Variables are stores in rows

```{r, warning=FALSE, message=FALSE, eval=FALSE}
weather <- tibble(
    station = c("Amsterdam", "Amsterdam", "Maastricht", "Maastricht"),
    element = c("min", "max", "min", "max"),
    d1 = c(4, 11, 3, 15),
    d2 = c(7, 10, 2, 10),
    d3 = c(5, 14, 5, 16)
)
weather
```

## Variables are stores in rows (the return of)

Now make wide dataset (`pivot_wider()`)

```{r, warning=FALSE, message=FALSE, eval=FALSE}
weather2 <- weather %>% 
    pivot_longer(
        d1:d3, 
        names_to = "day", 
        values_to = "temperature", 
        values_drop_na = TRUE
    ) %>% 
    pivot_wider(names_from = element, values_from = temperature)
```    

## Conditional logit example 

```{r, warning=FALSE, message=FALSE, eval=FALSE}
housing_prices <- tibble(
    prices = c(600000, 500000, 470000),
    city = c("Aadam", "Bedam", "Cedam"), 
    historic_centre = c(0, 0, 1)
)
housing_prices
``` 

## Conditional logit example (with a vengeance)

```{r, warning=FALSE, message=FALSE, eval=FALSE}
housing_prices <- housing_prices %>%
    mutate(obs = seq(1: nrow(housing_prices)))
    mutate(chosen = rep(1, nrow(housing_prices) ) ) %>%
    pivot_wider(names_from = city, 
                values_from = chosen) %>%
    replace(is.na(.), 0) %>%
    pivot_longer(Aadam:Cedam, 
                 names_to = "city",
                 values_to =  "chosen")
housing_prices
```

## In conclusion

- \alert{Script} everything you do with data

  - both ex-ante and ex-post analysis (during)
  
- `tidyverse` package provides excellent support

  - for small/medium sized datasets (until a couple of million obs.)
  
- tidying data (restructuring) is more important than you might think as you need to restructure some of your own data

  - \alert{avoid} using Excel
  - first \alert{think} about unit of analysis and what you want